---
title: "Methylation"
author: "Katherine I. Cooper"
date: "2026-05-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Methylation detection

Please run all chunks here and move to the parent file afterwards. This file contains **libraries** and a **function.**

```{r, echo = F}
#BioConductor is required for this. 
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# CRAN packages
base.pkgs <- c("tidyverse",
               "stringr")

for (pkg in base.pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

# Bioconductor packages
bioc.pkgs <- c("GenomicFeatures",
               "AnnotationDbi",
               "biomaRt",
               "org.Hs.eg.db",
               "org.Mm.eg.db",
               "BSgenome.Hsapiens.UCSC.hg38",
               "BSgenome.Mmusculus.UCSC.mm39"
               )

for (pkg in bioc.pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, update = TRUE, ask = FALSE, force = TRUE)
  }
  library(pkg, character.only = TRUE)
}


library(stringr)
library(GenomicFeatures)
library(BSgenome.Mmusculus.UCSC.mm39)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(TxDb.Mmusculus.UCSC.mm39.knownGene)
library(biomaRt)
```

## CG dinucleotide counter:

The following code cell contains a function that automatically compiles CG dinucleotides from a reference genome. To view the code, uncollapse the cell; however, this is non-essential for running the function.

To load the function, please click the green arrow on the right-side of the code cell. This will load the function into your environment. It is called: countCpG.


```{r, countCpG}

countCpG <- function(df, GeneIDs, flankRegion, species){
    
  # Set-up for the species specific portion
    
    #Which packages are needed
  message(paste("Loading packages for", species, "."))
  
  Species.pkgs <- switch(species,
    "Hs" =  c("BSgenome.Hsapiens.UCSC.hg38",
                   "org.Hs.eg.db"),
    "Mm" = c("BSgenome.Mmusculus.UCSC.mm39",
                        "org.Mm.eg.db"),
    "Gg" = c("BSgenome.Ggallus.UCSC.galGal5",
             "org.Gg.eg.db"),
    stop("Invalid species choice. Supported options: Homo sapiens (Hs), Mus musculus (Mm), Gallus gallus (Gg).")
    )
  
  sapply(Species.pkgs, library, character.only = T)
  
  #Assigning the ensembl 
  
   ensembl <- switch(species,
      "Hs" = useMart("ensembl", 
                     dataset = "hsapiens_gene_ensembl", 
                     host = "https://may2025.archive.ensembl.org"),
      "Mm" = useMart("ensembl", 
                     dataset = "mmusculus_gene_ensembl", 
                     host = "https://dec2021.archive.ensembl.org"),
      "Gg" = useMart("ensembl", 
                     dataset = "ggallus_gene_ensembl", 
                     host = "https://dec2021.archive.ensembl.org"),
      stop("Invalid species choice. Supported options: Homo sapiens (Hs), Mus musculus (Mm), Gallus gallus (Gg)"))
   
   BSgenome.sp <- switch(species,
     "Hs" = BSgenome.Hsapiens.UCSC.hg38,
     "Mm" = BSgenome.Mmusculus.UCSC.mm39,
     "Gg" = BSgenome.Ggallus.UCSC.galGal5,
     stop("Unknown BSgenome object")
   )
    
    genes = df[[GeneIDs]]
  
  #Mapping the ensembl names to the symbols
    
  #Getting which ensembl db
    
    pkg.map <- c(
      "Mm" = "org.Mm.eg.db",
      "Hs" = "org.Hs.eg.db",
      "Gg" = "org.Gg.eg.db"
    )
    
  org.db <- get(pkg.map[[species]])
  
  Ensembl_names <- mapIds(org.db,
    keys = genes, keytype = "SYMBOL", column = "ENSEMBL")
  
  ##loop set-ups 
  
  #Empty vector for the coords
  coords <- list()
  
  #Flanking regions
  flanks <- flankRegion
  
  #Counter for final df
  counter = 1
  
  
  ##Loop 1: Assigning TSS coordinate
  for (n in genes) {
    
    #Message for which gene
    
    message(paste0("(", counter, "/", length(Ensembl_names), ") Processing: ", n))
    
    #Checking for NA
    
    if(is.na(Ensembl_names[n]) || is.null(Ensembl_names[n])) {
      message(paste("Unknown gene SYMBOL ID for", n, "Skipping gene."))
      next
    }
    
    #Assinging the emsembl for the loop
    Ensembl_id <- Ensembl_names[n]
    
    #Getting the TSS coordinate
      tss_info <- getBM(
      filters = "ensembl_gene_id", 
      values = Ensembl_id,
      attributes = c("chromosome_name", "transcription_start_site", "strand"),
      mart = ensembl
    )
      
      #Skipping no TSS
      if(is.null(tss_info) || nrow(tss_info) == 0) {
        message(paste("Skipping gene:", n, "-No TSS information found in BioMart"))
        next
      }
      
      
      #Assigning the TSS
      
      tss_info <- tss_info[1, ]
    
      ##Loop 2: Finding CGs in the gene per each flanking regions
      for (i in flanks) {
    
      
        #Adding/subtracting flanks
    start_coord <- tss_info$transcription_start_site - i
    end_coord <- tss_info$transcription_start_site + i
    
    #Ensuring the start/stop are not flipped because of - strand
    if (start_coord > end_coord) {
      temp <- start_coord
      start_coord <- end_coord
      end_coord <- temp
    }
    
    #Pulling the DNA sequence info given the start/stop region
    tss_range <- GRanges(
      seqnames = paste0("chr", tss_info$chromosome_name),
      ranges = IRanges(start = start_coord, end = end_coord),
      strand = ifelse(tss_info$strand == 1, "+", "-")
    )
    
    #Getting the actual sequence
    tss_seqeunce <- getSeq(BSgenome.sp, tss_range)
    
    #Counting CG dinucleotides 
    CpG_count <- str_count(as.character(tss_seqeunce), "CG")
    
    #Data frame accumulation 
     coords[[counter]] <- data.frame(
       Gene_id = n,
       CpGs = CpG_count,
       flanking_region = as.character(i)
     )
    
     #Counter 
     counter <- counter + 1
     
     #Message for done
     
     message(paste("Finished processing", n, "."))
     
      }

  }
  cpgCounts <- do.call(rbind, coords)
  
  #Final message
  
  message("--- Processing complete!---")
  
  return(cpgCounts)
}

```

------------------------------------------------------------------------

#### Function documentation:

The following contains an example using a manually input data frame, genesTest. It is essential to have the proper syntax of the genes' symbols for mapping purpose: ex. RHO is incorrect in this instance. Rho is correct.

There are three arguments the user needs to input:

-   df: Identify the data containing the gene symbols

-   GeneIDs: Which column contains the gene symbol IDs in the specified df. This needs to be in quotation marks.

-   flankRegion: Specify the buffer regions around the TSS. This can be manually input or a vector.

**Common errors:** start_coord length is zero implies an improper gene name. I am working on making a more human-interpretable error code.

```{r}
sessionInfo()
```

