---
title: "Child_file"
output: pdf_document
---

# Child File

```{r set-up, include = FALSE}
knitr::set_parent("BIO611_parentFile.Rmd")
knitr::opts_chunk$set(fig.crop = FALSE)
```


The analyses in this file work off the Methylation parent file containing the countCpG function. Please run all chunks on that file prior to proceeding here.

```{r}
genes <- read.csv("C:/Users/kathe/Documents/Enke/Data/gene_cell-type_methylationInput.csv")

head(genes)
```

CpG counter

```{r, message=FALSE}
Mm_CpG <- countCpG(genes, "Gene", 
                   flankRegion = c(50,100,200,500,1000), 
                   species = "Mm")
```

```{r, message=FALSE}
#Conversting Gene ID to upper since human symbols use the full capital format.
genes$Gene <- toupper(genes$Gene)

Hs_CpG <- countCpG(genes, 
                   GeneIDs = "Gene", 
                   flankRegion = c(50, 100, 200, 500, 1000), 
                   species = "Hs")
```

### Plotting general distributions

#### Loading libraries 

These are all base CRAN libraries. Not Bioconductor.

```{r, message=F, results='hide'}
libs <- c("tidyverse",
          "car",
          "patchwork",
          "lme4",
          "ggeffects")

sapply(libs, library, character.only = T)
```

**Plotting per species**

Setting the theme. This is to ensure all the plots are cohesive with in the final product.
```{r, warning=FALSE, message=FALSE}

#Creating a theme for the document.
theme_childFile <- function(){
      theme(
    #Legend manipulation
    legend.position = "inside",
    legend.position.inside = c(0.1,0.8),
    legend.box.background = element_rect(color = "black",
                                linewidth = 1),
    legend.title = element_text(size = 14, 
                                face = "bold"),
    legend.text = element_text(size = 11,
                               color= "black"),
    #Panel manipulation
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black",
                                linewidth = 1),
    axis.ticks = element_line(linewidth = 1),
    axis.ticks.length = unit(5, "pt"),
    axis.minor.ticks.length = rel(0.5),
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(size = 15,
                              face = "bold"),
    #Text manipulation
    axis.title = element_text(face = "bold",
                              size = 17),
    axis.title.x = element_text(margin = margin(t = 7)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.text = element_text(size = 11, 
                             color = "black")
  )}
```


Mouse: 


```{r}
mouse.only <- ggplot() +
  geom_histogram(data = Mm_CpG, aes(CpGs), stat = "bin",
                 bins = 20,
                 fill = "darkgrey",
                 color = "black",
                 alpha = 0.5)  +
  labs(x = "Number of CpGs",
       y = "Count",
       title = "Mouse") +
  theme_bw() +
  theme_childFile()

human.only <- ggplot() +
  geom_histogram(data = Hs_CpG, aes(CpGs),
                 stat = "bin",
                 bins = 20,
                 fill ="orchid3",
                 color = "black",
                 alpha = 0.5) +
  labs(x = "Number of CpGs",
       y = "Count",
       title = "Human") +
  theme_bw() +
  theme_childFile()

combined <- ggplot() +
  geom_histogram(data = Mm_CpG, aes(CpGs), stat = "bin",
                 bins = 20,
                 fill = "darkgrey",
                 color = "black",
                 alpha = 0.5) +
  geom_histogram(data = Hs_CpG, aes(CpGs),
                 stat = "bin",
                 bins = 20,
                 fill = "orchid3",
                 color = "black",
                 alpha = 0.5)  +
  labs(x = "Number of CpGs",
       y = "Count") +
  theme_bw() +
  theme_childFile()

full.histograms <- (mouse.only | human.only)/combined

full.histograms
```


```{r}
#Creating CpG_total data frame to house the CpGs, organism, and cell-type
CpG_total <- Mm_CpG %>%
  mutate(
    Gene_id = toupper(Gene_id),
    organism = "Mm"
  ) %>%
  left_join(genes, by = join_by("Gene_id" == "Gene"))


#I did not do this in a great way. But it works.
Hs_temp <- Hs_CpG %>%
  mutate(organism = "Hs") %>%
  left_join(genes, by = join_by("Gene_id" == "Gene"))

CpG_total <- CpG_total %>%
  rbind(Hs_temp) %>%
  mutate(
    Cell.type = as.factor(Cell.type),
    organism = as.factor(organism),
    flanking_region = as.numeric(flanking_region),
    region = as.factor(flanking_region*2)
  )

rm(Hs_temp)
head(CpG_total)
```

### Visualizing both species distribution

Using a boxplot to look at the spread within cell-types and species.

```{r}
ggplot(CpG_total, aes(region, CpGs)) +
  
  #Adding in boxplot
  geom_boxplot(aes(fill = organism),
               staplewidth = 0.5,
               width = 0.5,
               alpha = 0.7,
               outlier.shape = 20,
               position = position_dodge(width = 0.7)) +
  
  #Adding in points
  geom_point(aes(color = organism),
             alpha = 0.5,
             shape = 18,
             position = position_dodge(width = 0.7)) +
  facet_wrap(~Cell.type, labeller = as_labeller(c(
    "both" = "Both",
    "rod" = "Rod only",
    "cone" = "Cone only"
  ))) +
  labs(x = "Flanking region around TSS",
       y = "Number of CG dinucleotides",
       fill = "Organism",
       color = "Organism") +
  guides(y = guide_axis(
    minor.ticks = TRUE
  )) +
  scale_fill_manual(values = c("grey", "orchid1"),
                    labels = c(
                      "Hs" = "Human",
                      "Mm" = "Mouse"
                    )) +
  scale_color_manual(values = c("grey1", "orchid4"),
                    labels = c(
                      "Hs" = "Human",
                      "Mm" = "Mouse"
                    )) +
  
  #Renaming since flanking region is a double and region was a factor. 
  scale_x_discrete(labels = c(
    "100" = "50",
    "200" = "100",
    "400" = "200",
    "1000" = "500",
    "2000" = "1000"
  )) +
  theme_bw() +
  theme_childFile()

#ggsave("CpG_Boxplots.png", height = 5, width = 7)
```


### Plotting each gene

Plotting to see the distribution within each gene

```{r, warning=FALSE, message=FALSE}
ggplot(CpG_total, aes(Gene_id, CpGs, color = organism)) +
  geom_point(position = position_dodge(width = 1),
             alpha = 0.3) +
  stat_summary(fun = "mean", 
               position = position_dodge(width = 1)) +
  stat_summary(fun = "mean",
               geom = "point",
               shape = 95,
               size = 8,
               position = position_dodge(width=1)) +
  facet_wrap(~Cell.type, 
             nrow = 3,
             scales = "free", 
             labeller = as_labeller(c(
                "both" = "Both",
                "rod" = "Rod only",
                "cone" = "Cone only")))  +
  scale_color_manual(values = c("darkgrey", "orchid"),
                    labels = c(
                      "Hs" = "Human",
                      "Mm" = "Mouse"
                    )) +
  labs(y = "Mean CpG count",
       x = "Gene ID",
       color = "Organism") +
  guides(y = guide_axis(minor.ticks = TRUE)) +
  theme_bw() +
  theme(
    #Legend manipulation
    legend.position = "top",
    legend.title = element_text(size = 14, 
                                face = "bold"),
    legend.text = element_text(size = 11,
                               color= "black"),
    #Panel manipulation
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black",
                                linewidth = 1),
    axis.ticks = element_line(linewidth = 1),
    axis.ticks.length = unit(5, "pt"),
    axis.minor.ticks.length = rel(0.5),
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(size = 15,
                              face = "bold"),
    #Text manipulation
    axis.title = element_text(face = "bold",
                              size = 17),
    axis.title.x = element_text(margin = margin(t = 7)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.text = element_text(size = 11, 
                             color = "black")
  )


#ggsave("CpG_GenePlot.png", height = 9, width = 9)
```

---

### Data analysis

I am doing a GLM with a poisson model and an ANOVA with standardized CpG counts. This is due to the left-skewed data as well as it being count data with a high dispersion. 

**GLM** 

```{r}
cpg.glm <- glmer.nb(CpGs ~ Cell.type * organism  + (1|Gene_id), 
                 offset = log(flanking_region *2),
                 data = CpG_total)
                 
 summary(cpg.glm)
```

### Plotting GLMM 

```{r, predictingChunk}
#Creating a predicted response for the GLMM 
predict.2 <- predict_response(cpg.glm, terms = c("Cell.type", "organism"),
                              condition = list(flanking_region = 0.5))

predict.2
```

#### Plotting

Predicting the number of CpGs using the NB GLMM created. 

```{r, fig.align='left'}
#Plotting the predicted variables

predict.Plot <- ggplot(predict.2, aes(x, predicted, color = group)) +
  geom_point(position = position_dodge(width = 0.5),
             size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                position = position_dodge(width = 0.5),
                width = 0.2) +
  scale_x_discrete(labels = c(
    both = "Both",
    cone = "Cone-only",
    rod = "Rod-only"
  )) +
  scale_color_manual(values = c("darkgrey", "orchid"),
                    labels = c(
                      "Hs" = "Human",
                      "Mm" = "Mouse"
                    )) +
  labs(y = "Predicted response",
       x = "Cell type",
       color = "Species") +
  guides(y = guide_axis(minor.ticks = TRUE)) +
  theme_bw()+
  theme_childFile() +
  theme(legend.position = "top",
        panel.grid.major.x = element_line(linetype = "dashed"),
        legend.background = element_rect(color = "white", 
                                         linewidth = 0))

predict.Plot

#ggsave("C:/Users/kathe/Downloads/GenePlot.png")
```


#### Creating a plot with the observed means

```{r, fig.align='left'}
#GLMM uses an offset with the flanking region which needs to be accounted for when plotting. 
CpG.diff <- CpG_total %>%
  mutate(
    standardized = CpGs/(flanking_region * 2)
  )

#Creating the observed plot
observed.Plot <- ggplot(CpG.diff, aes(x = Cell.type, y = standardized, color = organism)) +
  stat_summary(fun = "mean", geom = "point",
               position = position_dodge(width = 0.5)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar",
               position = position_dodge(width = 0.5),
               width = 0.2) +
  scale_x_discrete(labels = c(
    both = "Both",
    cone = "Cone-only",
    rod = "Rod-only"
  )) +
  scale_color_manual(values = c("darkgrey", "orchid"),
                    labels = c(
                      "Hs" = "Human",
                      "Mm" = "Mouse"
                    )) +
  labs(y = "Observed values",
       x = "Cell type",
       color = "Species") +
  guides(y = guide_axis(minor.ticks = TRUE),
         color = "none") +
  theme_bw()+
  theme_childFile() +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(linetype = "dashed"),
        legend.background = element_rect(color = "white", 
                                         linewidth = 0))

observed.Plot
```

#### Combining plots for a full figure.

```{r, fig.align='left'}
full.plot <- observed.Plot | predict.Plot

full.plot + plot_layout(guides = "collect") &
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag = element_text(face = "bold",
                                size = 17)) 

#ggsave("GLMM_CpG.plot.png", width = 8, height = 6)
```

#### Checking assumptions for GLMM

Checking assumptions of the dispersion.

```{r}
library(DHARMa)
sim_res <- simulateResiduals(cpg.glm)
plot(sim_res)
testDispersion(sim_res)
```

Checking the assumptions for the rest of it.

```{r}
plot(log(CpG_total$flanking_region * 2), log(CpG_total$CpGs + 0.5))
#Should show roughly linear trend with slope ~1
library(lattice)
dotplot(ranef(cpg.glm, condVar = TRUE))
#Points should be roughly symmetrically distributed around 0
```


Looking at the exponential version of the coeffecients for interpretation. The coefficents are in log format and need to be converted for proper analysis.

```{r}
exp(coef(summary(cpg.glm)))
```




