---
title: "Genomic plots"
author: "Katherine I. Cooper"
date: "2026-05-06"
output:
  pdf_document: default
  html_document: default
---
# Genomic plots for comparison

A few genes have vastly different mean CpG counts: _NRL, THRB_.

**GOAL:** Look at the sequences for CpG islands

```{r, include=FALSE}

# CRAN packages
cran.pkgs <- c("tidyverse",
               "patchwork",
               "gridExtra",
               "BiocManager"
                )
for (pkg in cran.pkgs) {
   if (!requireNamespace(pkg, quietly = TRUE)) {
     install.packages(pkg, update = TRUE, ask = FALSE, force = TRUE)
   }
   library(pkg, character.only = TRUE)
 }


# Bioconductor packages
bioc.pkgs <- c("GenomicFeatures",
              "AnnotationDbi",
              "biomaRt",
              "org.Hs.eg.db",
              "org.Mm.eg.db",
              "org.Gg.eg.db",
              "BSgenome.Hsapiens.UCSC.hg38",
              "BSgenome.Mmusculus.UCSC.mm39",
              "ggbio"
                )
for (pkg in bioc.pkgs) {
   if (!requireNamespace(pkg, quietly = TRUE)) {
     BiocManager::install(pkg, update = TRUE, ask = FALSE, force = TRUE)
   }
   library(pkg, character.only = TRUE)
 }

```

### Step 1: Getting the sequences for 1,000 bp

```{r}
#Creating a dataframe of the genes
Genes.of.int <- data.frame (
  "Human" = c("THRB", "NRL"),
  "Mouse" = c("Thrb", "Nrl"))

#Mapping the gene symbols to the Ensembl
Ensembl_names.Hs <- mapIds(org.Hs.eg.db, 
                        keys = Genes.of.int$Human, 
                        keytype = "SYMBOL", 
                        column = "ENSEMBL")

Ensembl_names.Mm <- mapIds(org.Mm.eg.db,
                           keys = Genes.of.int$Mouse,
                           keytype = "SYMBOL",
                           column = "ENSEMBL")

#Adding to the Df. 
Genes.of.int$Ensembl.Mm <- Ensembl_names.Mm
Genes.of.int$Ensembl.Hs <- Ensembl_names.Hs

head(Genes.of.int)
```

I will be using the 1kb flanking region around the canonical start site. 

#### Starting with human 

```{r}
#Using Ensembl mart archive for a stable connection:
ensembl.hs <- useMart("ensembl",
                      dataset = "hsapiens_gene_ensembl",
                      host = "https://may2025.archive.ensembl.org")

#Acquiring the TSS info
tss_info.Hs.THRB <- getBM(filters = "ensembl_gene_id", 
                  values = Genes.of.int$Ensembl.Hs[1], 
                  attributes = c("chromosome_name", "transcription_start_site", "strand"), 
                  mart = ensembl.hs)

#Getting the canonical start for THRB
tss_info.Hs.THRB <- tss_info.Hs.THRB[1, ]

#Creating the range...THRB seems to be reverse so the start and end are flipped.
start <- tss_info.Hs.THRB$transcription_start_site +1000 
end <- tss_info.Hs.THRB$transcription_start_site - 1000

#Flipping start and end...
tss_range.Hs.THRB <- GRanges(seqnames = paste0("chr", tss_info.Hs.THRB$chromosome_name), 
                       ranges = IRanges(start = end, end = start), 
                       strand = ifelse(tss_info.Hs.THRB$strand == 1, "+", "-"))

remove(start)
remove(end)
head(tss_range.Hs.THRB)
```

Getting the BS DNA 

```{r}
DNA.Hs.THRB <- getSeq(BSgenome.Hsapiens.UCSC.hg38, tss_range.Hs.THRB)

head(DNA.Hs.THRB)
```

#### Repeating for the other four... 

I do not want to make a loop. I am sorry :(

```{r}
#Acquiring the TSS info
tss_info.Hs.NRL <- getBM(filters = "ensembl_gene_id", 
                  values = Genes.of.int$Ensembl.Hs[2], 
                  attributes = c("chromosome_name", "transcription_start_site", "strand"), 
                  mart = ensembl.hs)

#Getting the canonical start for NRL
tss_info.Hs.NRL <- tss_info.Hs.NRL[1, ]

#Creating the range...THRB seems to be reverse so the start and end are flipped.
start <- tss_info.Hs.NRL$transcription_start_site +1000 
end <- tss_info.Hs.NRL$transcription_start_site - 1000

#Flipping start and end...
tss_range.Hs.NRL <- GRanges(seqnames = paste0("chr", tss_info.Hs.NRL$chromosome_name), 
                       ranges = IRanges(start = end, end = start), 
                       strand = ifelse(tss_info.Hs.NRL$strand == 1, "+", "-"))

remove(start)
remove(end)

DNA.Hs.NRL <- getSeq(BSgenome.Hsapiens.UCSC.hg38, tss_range.Hs.NRL)
head(DNA.Hs.NRL)
```

#### Now for mouse!

```{r}
ensembl.mm <- useMart("ensembl", 
                     dataset = "mmusculus_gene_ensembl", 
                     host = "https://dec2021.archive.ensembl.org")

#Acquiring the TSS info
tss_info.Mm.Thrb <- getBM(filters = "ensembl_gene_id", 
                  values = Genes.of.int$Ensembl.Mm[1], 
                  attributes = c("chromosome_name", "transcription_start_site", "strand"), 
                  mart = ensembl.mm)

#Getting the canonical start for THRB
tss_info.Mm.Thrb <- tss_info.Mm.Thrb[1, ]

#Creating the range...THRB seems to be reverse so the start and end are flipped.
start <- tss_info.Mm.Thrb$transcription_start_site +1000 
end <- tss_info.Mm.Thrb$transcription_start_site - 1000

#Flipping start and end...
tss_range.Mm.Thrb <- GRanges(seqnames = paste0("chr", tss_info.Mm.Thrb$chromosome_name), 
                       ranges = IRanges(start = end, end = start), 
                       strand = ifelse(tss_info.Mm.Thrb$strand == 1, "+", "-"))

remove(start)
remove(end)

DNA.Mm.Thrb <- getSeq(BSgenome.Mmusculus.UCSC.mm39, tss_range.Mm.Thrb)
head(DNA.Mm.Thrb)
```

NRL
```{r}
#Acquiring the TSS info
tss_info.Mm.Nrl <- getBM(filters = "ensembl_gene_id", 
                  values = Genes.of.int$Ensembl.Mm[2], 
                  attributes = c("chromosome_name", "transcription_start_site", "strand"), 
                  mart = ensembl.mm)

#Getting the canonical start for NRL
tss_info.Mm.Nrl <- tss_info.Mm.Nrl[1, ]

#Creating the range...THRB seems to be reverse so the start and end are flipped.
start <- tss_info.Mm.Nrl$transcription_start_site +1000 
end <- tss_info.Mm.Nrl$transcription_start_site - 1000

#Flipping start and end...
tss_range.Mm.Nrl <- GRanges(seqnames = paste0("chr", tss_info.Mm.Nrl$chromosome_name), 
                       ranges = IRanges(start = end, end = start), 
                       strand = ifelse(tss_info.Mm.Nrl$strand == 1, "+", "-"))

remove(start)
remove(end)

DNA.Mm.Nrl <- getSeq(BSgenome.Mmusculus.UCSC.mm39, tss_range.Mm.Nrl)
head(DNA.Mm.Nrl)
```

### Step 2: plotting using ggBio 

Need to convert the sequences into characters to parse for any occurance of CG. 

#### Hs NRL

```{r}

#Convert sequence to a character vector
dna_str <- as.character(DNA.Hs.NRL)
bases <- strsplit(dna_str, "")[[1]]

#Create a status vector: default to "Other" to separate out the "CG"s
color_status <- rep("Other", length(bases))

#Identify CpGs
for (i in 1:(length(bases) - 1)) {
  if (bases[i] == "C" && bases[i+1] == "G") {
    color_status[i] <- "CG"
    color_status[i+1] <- "CG"
  }
}

#Expand your TSS range into 1-bp intervals
gr_expanded.HN <- unlist(tile(tss_range.Hs.NRL, width = 1))

#Add the new metadata column
mcols(gr_expanded.HN)$is_cg <- color_status

Hs.NRL.plot <- autoplot(gr_expanded.HN, geom = "rect", aes(fill = is_cg, color = is_cg)) +
  geom_vline(aes(xintercept = tss_info.Hs.NRL$transcription_start_site), color = "black", linewidth = 1.5) +
  scale_fill_manual(values = c("CG" = "#E41A1C",
                               "Other" = "#D3D3D3")) + 
  scale_color_manual(values = c("CG" = "#E41A1C",  
                               "Other" = "#D3D3D3")) +
  labs(color = "CpG",
       fill = "CpG",
       title = "Human NRL") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", 
                              size = 20,
                              hjust = 0.5,
                              margin = margin(b = 15, unit = "pt")),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 17,
                                face = "bold"),
    legend.position = "top"
  )

Hs.NRL.plot
```

#### Hs THRB

```{r, HsTHRBPlot}

#Convert sequence to a character vector
dna_str <- as.character(DNA.Hs.THRB)
bases <- strsplit(dna_str, "")[[1]]

#Create a status vector
color_status <- rep("Other", length(bases))

# Identify CpGs
for (i in 1:(length(bases) - 1)) {
  if (bases[i] == "C" && bases[i+1] == "G") {
    color_status[i] <- "CG"
    color_status[i+1] <- "CG"
  }
}

#Expand your TSS range into 1-bp intervals
gr_expanded.HT <- unlist(tile(tss_range.Hs.THRB, width = 1))

#Add the new metadata column
mcols(gr_expanded.HT)$is_cg <- color_status

#Autoplot using ggbio
Hs.THRB.plot <- autoplot(gr_expanded.HT, geom = "rect", aes(fill = is_cg, color = is_cg)) +
  geom_vline(aes(xintercept = tss_info.Hs.THRB$transcription_start_site), color = "black", linewidth = 1.5) +
  scale_fill_manual(values = c("CG" = "#E41A1C",
                               "Other" = "#D3D3D3")) + 
  scale_color_manual(values = c("CG" = "#E41A1C",  
                               "Other" = "#D3D3D3")) +
  labs(color = "CpG",
       fill = "CpG",
       title = "Human THRB") +
  
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", 
                              size = 20,
                              hjust = 0.5,
                              margin = margin(b = 15, unit = "pt")),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 17,
                                face = "bold"),
    legend.position = "none"
  )


Hs.THRB.plot

```

#### Mm NRL

```{r, MmNRLPlot}

#Convert sequence to a character vector
dna_str <- as.character(DNA.Mm.Nrl)
bases <- strsplit(dna_str, "")[[1]]

#Create a status vector
color_status <- rep("Other", length(bases))

#Identify CpGs
for (i in 1:(length(bases) - 1)) {
  if (bases[i] == "C" && bases[i+1] == "G") {
    color_status[i] <- "CG"
    color_status[i+1] <- "CG"
  }
}

#Expand your TSS range into 1-bp intervals
gr_expanded.MN <- unlist(tile(tss_range.Mm.Nrl, width = 1))

#Add the new metadata column
mcols(gr_expanded.MN)$is_cg <- color_status

Mm.Nrl.plot <- autoplot(gr_expanded.MN, geom = "rect", aes(fill = is_cg, color = is_cg)) +
  geom_vline(aes(xintercept = tss_info.Mm.Nrl$transcription_start_site), color = "black", linewidth = 1.5) +
  scale_fill_manual(values = c("CG" = "#E41A1C",
                               "Other" = "#D3D3D3")) + 
  scale_color_manual(values = c("CG" = "#E41A1C",  
                               "Other" = "#D3D3D3")) +
  labs(color = "CpG",
       fill = "CpG",
       title = "Mouse NRL") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", 
                              size = 20,
                              hjust = 0.5,
                              margin = margin(b = 15, unit = "pt")),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 17,
                                face = "bold"),
    legend.position = "none"
  )


Mm.Nrl.plot
```

#### Mm THRB

```{r, MmTHRBPlot}

#Convert sequence to a character vector
dna_str <- as.character(DNA.Mm.Thrb)
bases <- strsplit(dna_str, "")[[1]]

#Create a status vector
color_status <- rep("Other", length(bases))

#Identify CpGs
for (i in 1:(length(bases) - 1)) {
  if (bases[i] == "C" && bases[i+1] == "G") {
    color_status[i] <- "CG"
    color_status[i+1] <- "CG"
  }
}

#Expand your TSS range into 1-bp intervals
gr_expanded.MT <- unlist(tile(tss_range.Mm.Thrb, width = 1))

#Add the new metadata column
mcols(gr_expanded.MT)$is_cg <- color_status

Mm.Thrb.plot <- autoplot(gr_expanded.MT, geom = "rect", aes(fill = is_cg, color = is_cg)) +
  geom_vline(aes(xintercept = tss_info.Mm.Thrb$transcription_start_site), color = "black", linewidth = 1.5) +
  scale_fill_manual(values = c("CG" = "#E41A1C",
                               "Other" = "#D3D3D3")) + 
  scale_color_manual(values = c("CG" = "#E41A1C",  
                               "Other" = "#D3D3D3")) +
  labs(color = "CpG",
       fill = "CpG",
       title = "Mouse THRB") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", 
                              size = 20,
                              hjust = 0.5,
                              margin = margin(b = 15, unit = "pt")),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 17,
                                face = "bold"),
    legend.position = "none"
  )

Mm.Thrb.plot
```

### Step 3: Combining all

Combining all the plots using gridExtra to account for the ggbio objects.

```{r, finalplot, fig.height=8, fig.width=4}

#Converting ggbio objects to ggplot to use with gridExtra
p1 <- Hs.NRL.plot@ggplot
p2 <- Mm.Nrl.plot@ggplot
p3 <- Hs.THRB.plot@ggplot
p4 <- Mm.Thrb.plot@ggplot

#Combine with collected legends so the sizes are all the same
full.cpg<- wrap_plots(p1, p2, p3, p4, ncol = 1,  heights = c(2, 2, 2, 2)) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")

full.cpg
```

```{r}
sessionInfo()
```








